---
title: "Week4_Notes"
author: "Alexander Frieden"
date: "February 13, 2016"
output: ioslides_presentation
---

## PCA

* This week we will work with a statistical tool called principal component analysis.  

## What is it?

is a statistical procedure that uses an orthogonal transformation to convert a set of observations of possibly correlated variables into a set of values of linearly uncorrelated variables called principal components. 

Before we get into it, lets introduce the various topics

## Covariance

* Standard deviation and variance only operate on 1 dimension, so that you could
only calculate the standard deviation for each dimension of the data set independently of the other dimensions. 

* However, it is useful to have a similar measure to find out how
much the dimensions vary from the mean with respect to each other

* Covariance is such a measure. Covariance is always measured between 2 dimensions.  If you calculate the covariance between one dimension and itself, you get the variance. 

## Covariance (part 2)

* So, if you had a 3-dimensional data set $(x, y, z)$, then you could measure the
covariance between the x and y dimensions, the x and z dimensions, and the y and z
dimensions. Measuring the covariance between x and x, or y and y, or z and z would
give you the variance of the x, y and z dimensions respectively.

## Covariance (part 3)

The formula for covariance is very similar to the formula for variance. The formula
for variance could also be written like this: 
$$
var(X) = \frac{\sum_{i=1}^n(X_i-\bar{X})(X_i-\bar{X})}{n-1}
$$

where I have simply expanded the square term to show both parts. So given that knowledge,
here is the formula for covariance:

$$
COV(X) = \frac{\sum_{i=1}^n(X_i-\bar{X})(Y_i-\bar{Y})}{n-1}
$$

## Covariance Matrix

A useful way to get all the possible covariance values between all the different
dimensions is to calculate them all and put them in a matrix. So, the definition for
the covariance matrix for a set of data with $n$ dimensions is:
$$
C^{n\times n} = (c_{ij},c_{ij} = cov(DIM_i,DIM_j))
$$

where $C^{n\times n}$ is a matrix with $n$ rows and $n$ columns, and $DIM_x$ is the $x^{th}$ dimension.

## Covariance Matrix Example

3 dimensional covariance matrix
$$
C = \left( \begin{array}{ccc}
cov(x,x) & cov(x,y) & cov(x,z) \\
cov(y,x) & cov(y,y) & cov(y,z) \\
cov(z,x) & cov(z,y) & cov(z,z) \end{array} \right)
$$

Notice the matrix is symmetrical about the diagonal.  Why?
<div class="notes">
Because $cov(a,b) = cov(b,a)$
</div>

## Covariance Example

Calculate covariance between x and y

      x                     y     
------------           ----------   
     10                   43        
     39                   13      
     19                   32
     23                   21
     28                   20

## Covariance Example (part 2)
Mean of x = 23.8, Mean of y = 25.8
$$
\begin{align}
COV(X) & = \frac{\sum_{i=1}^n(X_i-\bar{X})(Y_i-\bar{Y})}{n-1} \\
& = \frac{(-13.8)(17.2) + (15.2)(-12.8) + (-4.8)(6.2) + (-0.8)(-4.8) + (4.2)(-5.8)}{5-1} \\
& = \frac{-237.36 + -194.56 + -29.76 + 3.84 + -24.36}{4} \\
& = \frac{-482.2}{4} = -120.55
\end{align}
$$

## Covariance Example (part 3)

or we can do 
```{r covariance_example}
x=c(10,39,19,23,28)
y=c(43,13,32,21,20)
cov(x,y)
```

## Matrix Multiplication
Example of one non-eigenvector and one eigenvector.  

$$
 \left( \begin{array}{cc}
2 & 3 \\
2 & 1
\end{array} \right) \times \left( \begin{array}{cc}
1 \\
3
\end{array} \right) =\left( \begin{array}{cc}
11 \\
5
\end{array} \right)
$$
and
$$
 \left( \begin{array}{cc}
2 & 3 \\
2 & 1
\end{array} \right) \times \left( \begin{array}{cc}
3 \\
2
\end{array} \right) =\left( \begin{array}{cc}
12 \\
8
\end{array} \right) =4\left( \begin{array}{cc}
3 \\
2
\end{array} \right)
$$

## Matrix Multiplication (part 2)
How a scaled eigenvector is still and eigenvector
$$
2 \times \left( \begin{array}{cc}
3 \\
2
\end{array} \right) =\left( \begin{array}{cc}
6 \\
4
\end{array} \right)
$$
$$
 \left( \begin{array}{cc}
2 & 3 \\
2 & 1
\end{array} \right) \times \left( \begin{array}{cc}
6 \\
4
\end{array} \right) =\left( \begin{array}{cc}
24 \\
16
\end{array} \right) = 4\left( \begin{array}{cc}
6 \\
4
\end{array} \right)
$$

## Eigenvectors

* As you know, you can multiply two matrices together, provided they are compatible
sizes. 

* Eigenvectors are a special case of this. Consider the two multiplications between
a matrix and a vector in the first matrix example

## Eigenvectors (part 2)

* In the first example, the resulting vector is not an integer multiple of the original
vector, whereas in the second example, the example is exactly 4 times the vector we
began with. 

* Why is this? Well, the vector is a vector in 2 dimensional space. The vector [3 2] (from the second example multiplication) represents an arrow pointing from the origin.  The other matrix, the square one, can be thought of as a transformation matrix.

## Properties of Eigenvectors

You should first know that eigenvectors
can only be found for square matrices. And, not every square matrix has eigenvectors.
And, given an $n\times n$ matrix that does have eigenvectors, there are $n$ of them.
Given a $3\times 3$ matrix, there are 3 eigenvectors.

Another property of eigenvectors is that even if I scale the vector by some amount
before I multiply it, I still get the same multiple of it as a result.  

Lastly, all the eigenvectors of a matrix are perpendicular,
ie. at right angles to each other, no matter how many dimensions you have i.e. orthogonal.

## Eigenvectors (part 3)

* Lastly, all the eigenvectors of a matrix are perpendicular,
ie. at right angles to each other, no matter how many dimensions you have i.e. orthogonal.

* This is important because it means that you can express the data in terms of these perpendicular eigenvectors,
instead of expressing them in terms of the $x$ and $y$ axes. We will be doing this later in the section on PCA.

* Often times, eigenvectors are normalized so their length is one.  

## Finding Eigenvectors

How does one go about finding these mystical eigenvectors? Unfortunately, it’s
only easy(ish) if you have a rather small matrix, like no bigger than about $3\times 3$.  

* The usual way to find the eigenvectors is by complicated iterative methods, which are sometimes neccesary.  

## Eigenvalues

In the previous examples, we saw we got an eigenvector of 4.  

Eigenvectors and eigenvalues always come in pairs.  

## Principal Components Analysis

* It is a way of identifying patterns in data, and expressing the data in such a way as to highlight
their similarities and differences

* Since patterns in data can be hard to find in data of high dimension, where the luxury of graphical representation is not available, PCA is a powerful tool for analysing data


## PCA Method Part 1: Substract the mean

For PCA to work properly, you have to subtract the mean from each of the data dimensions.
The mean subtracted is the average across each dimension. So, all the $x$ values have $\bar{x}$ (the mean of $x$) subtracted and all the $y$ values subtracted from $\bar{y}$.  

This produces a data set whose mean is zero.




## VCF File

Variant Call Format

## Install Bioconductor and packages

```{r install_packages}
source("http://bioconductor.org/biocLite.R")
biocLite("gdsfmt")
biocLite("SNPRelate")
# Load the R packages: gdsfmt and SNPRelate
library(gdsfmt)
library(SNPRelate)
```

## PCA on vcf data (part 1)
```{r load_vcf_data}
## Load our vcf data
vcf.fn <- system.file("extdata", "sequence.vcf", package="SNPRelate")
## we can also explictly load the data
## vcf.fn <- "C:/your_folder/your_vcf_file.vcf"
```

## PCA on vcf data (part 2)

```{r}
# Reformat
snpgdsVCF2GDS(vcf.fn, "test.gds", method="biallelic.only")
snpgdsSummary("test.gds")
```

## PCA on vcf data (part 3)
```{r}
# Open the GDS file
genofile <- snpgdsOpen(snpgdsExampleFileName())
pop_code <- read.gdsn(index.gdsn(genofile, path="sample.annot/pop.group"))
table(pop_code)
```


## PCA on vcf data (part 4)

```{r hap_map_ethnicity_values}
# Display the first six values
head(pop_code)
```

## HapMap populations

    Key     Description    
-------     ------ 
    ASW     African ancestry in Southwest USA    
    CEU     Utah residents with Northern and Western European ancestry from the CEPH collection   
    CHB     Han Chinese in Beijing, China
    CHD     Chinese in Metropolitan Denver, Colorado
    GIH     Gujarati Indians in Houston, Texas
    JPT     Japanese in Tokyo, Japan
    LWK     Luhya in Webuye, Kenya
    MXL     Mexican ancestry in Los Angeles, California

## HapMap populations (part 2)
    Key     Description  
-------     ------ 
    MKK     Maasai in Kinyawa, Kenya
    TSI     Toscani in Italia
    YRI     Yoruba in Ibadan, Nigeria
    
## PCA on vcf data (part 5)

* It is suggested to use a pruned set of SNPs which are in approximate linkage equilibrium with each other to avoid the strong influence of SNP clusters in principal component analysis and relatedness analysis.

* Interesting paper relating to this here:
https://cdr.lib.unc.edu/indexablecontent/uuid:0a68bb86-def4-4493-ae41-2d3228668b84

* This is called LD-based SNP pruning.  

## PCA on vcf data (part 6)

```{r ld_snp_pruning}
set.seed(1000)
# Try different LD thresholds for sensitivity analysis
snpset <- snpgdsLDpruning(genofile, ld.threshold=0.2)
```


## PCA on vcf data (part 7)
Now we can see all the names of our snpset.  Question: what do we not see here?

```{r}
names(snpset)
# Get all selected snp id
snpset.id <- unlist(snpset)
```


## PCA on vcf data (part 8)
The functions in SNPRelate for PCA include calculating the genetic covariance matrix from genotypes, computing the correlation coefficients between sample loadings and genotypes for each SNP, calculating SNP eigenvectors (loadings), and estimating the sample loadings of a new dataset from specified SNP eigenvectors.

## Quick review: Covariance matrix and covariance

## Review eigenvectors and eigenvalues

## PCA on vcf data (part 9)

```{r run_pca}
# Run PCA
pca <- snpgdsPCA(genofile, snp.id=snpset.id, num.thread=2)
```

## PCA on vcf data (part 10)

* The following codeshows how to calculate the percent of variation is accounted for by the top principal components. 

* It is clear to see the first two eigenvectors hold the largest percentage of variance among the population, although the total variance accounted for is still less the one-quarter of the total.

## PCA on vcf data (part 11)

```{r}
# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))
```

## PCA on vcf data (part 12)
In the case of no prior population information, we can compose it
```{r}
tab <- data.frame(sample.id = pca$sample.id,
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    stringsAsFactors = FALSE)
head(tab)
```

## PCA on vcf data (part 13)

```{r plot_pca}
# Draw
plot(tab$EV2, tab$EV1, xlab="eigenvector 2", ylab="eigenvector 1")
```

## PCA on vcf data (part 14)

```{r}
# Get sample id
sample.id <- read.gdsn(index.gdsn(genofile, "sample.id"))
# Get population information
#   or pop_code <- scan("pop.txt", what=character())
#   if it is stored in a text file "pop.txt"
pop_code <- read.gdsn(index.gdsn(genofile, "sample.annot/pop.group"))
# assume the order of sample IDs is as the same as population codes
head(cbind(sample.id, pop_code))
```

## PCA on vcf data (part 15)
```{r}
# Make a data.frame
tab <- data.frame(sample.id = pca$sample.id,
    pop = factor(pop_code)[match(pca$sample.id, sample.id)],
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    stringsAsFactors = FALSE)
head(tab)
```


## PCA on vcf data (part 16)
```{r}
# Draw
plot(tab$EV2, tab$EV1, col=as.integer(tab$pop), xlab="eigenvector 2", ylab="eigenvector 1")
legend("bottomright", legend=levels(tab$pop), pch="o", col=1:nlevels(tab$pop))
```